### 2016-03-08 paper/icfp16-submission/semantics.tex

\paragraph{Inputs and Outputs}
%
The problem of generating inputs is undecidable in general.
%
Our witness generation procedure takes two inputs.
%
First, a search bound $n$ which is used to define the \emph{number} of
traces to explore. (We assume, without loss of generality, that all
traces are finite.)
%
Second, the target expression $e$ that contains the type error,
and may be a curried function of multiple arguments.
%
The witness generation procedure returns as output a list of (general)
witness expressions, each of which is of the form $e\ v_1 \ldots v_n$.
%
The \emph{empty} list is returned when no witness can be found after
exploring $n$ traces.

\paragraph{Modeling Semantics}
%
We resolve the non-determinism in the operational semantics
(\S~\ref{sec:semantics}) via the procedure
%
$$
\evalN : e \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu}^{*}
$$
%
Due to the non-determinism introduced by \gensym, a call
$\evalfn{e}$ returns the \emph{list}
of possible results of the form $\triple{v \cup \stuck}{\vsu}{\tsu}$
such that $\steps{e}{\emptysu}{\emptysu}{v \cup \stuck}{\vsu}{\tsu}$.

We address the issue of currying by defining a procedure \genArgs{e},
defined in Figure~\ref{fig:algo-gen-args},
that yields a list of holes $v_1, \ldots, v_n$ such that
$\eapp{e}{v_1 \ldots v_n}$ \emph{does not} evaluate to a lambda.
%
This is achieved via a loop that keeps adding holes to the
target application until evaluating the term yields a
non-lambda value.
%
% The helper functions @witness@ and @close@ are used to respectively
% apply the parameters to the target and close the result under the top-level
% let-binders, prior to invoking \hbox{@eval@.}
%
% That is, @mkApps@ creates a nested sequence of applications in
% the usual left-associative style, and @mkLets@ takes a list of
% binders and a body expression, and creates a sequence of nested
% let-binders that close the body expression.

\begin{figure}[t]
% $$
% \begin{array}{lcl}
% \genArgs{e} &\defeq& \begin{cases} \genArgs{\addArg{e}},& \mbox{if } \triple{\efun{x}{e'}}{\vsu}{\tsu},\ldots = \doeval{e} \\
%  e,& \mbox{otherwise} \\
% \end{cases}
% \\
% \addArg{e} &\defeq& \eapp{e}{\vhole{\thole}}, \qquad \mbox{where } \ehole, \thole \mbox{ are fresh }
% \end{array}
% $$
$$
\begin{array}{lcl}
\genArgsN   & : & e \rightarrow e^{*} \\
\genArgs{e} & = & \loopfn{[]} \\
\quad \mbox{\textbf{where}} &    & \\
  \quad \quad \loopfn{vs} & = & \mbox{\textbf{case }} \evalfn{\eapp{e}{vs}} \mbox{\textbf{ of}} \\
             &   & \quad \triple{\efun{x}{e}}{\vsu}{\tsu},\ldots \rightarrow \loopfn{\vhole{\thole} :: vs} \\
             &   & \qquad (\ehole, \thole \mbox{ are fresh}) \\
             &   & \quad \_ \rightarrow vs
\end{array}
$$

% \begin{mcode}
% genArgs :: $e$ -> [$v$]
% genArgs $e$ = loop []
%   where
%   loop vs = case eval $\triple{\eapp{e}{vs}}{\emptysu}{\emptysu}$ of
%              $\triple{\efun{x}{e}}{\vsu}{\tsu}$:_ -> loop ($\vhole{\thole}$ : vs)
%                            -- $\ehole$, $\thole$ are fresh
%              _          -> vs
% \end{mcode}
\caption{Finding the number of arguments for the target expression.}
\label{fig:algo-gen-args}
\end{figure}

\paragraph{Generating Witnesses}
%
Finally, Figure~\ref{fig:algo-gen-witness} summarizes the overall
implementation of our search for witnesses, in the procedure
$\genWitness{k}{e}$, which takes as input a bound $k$ and the
target expression $e$, and returns a list of witness expressions
$\eapp{e}{v_1 \ldots v_n}$ that demonstrate how the input program
gets stuck.
%
The search proceeds as follows.
%
\begin{enumerate}
  \item We invoke $\genArgsN$ to determine the number of
        parameters required by the (curried) target expression.

  \item We take the first $k$ traces returned by $\evalN$
        on the target $\eapp{e}{vs}$, and

@@

\begin{cor}{\textbf{[Witness Generation]}}
\label{thm:generation}
  If $\genWitness{k}{e} = \triple{\eapp{e}{v_1 \ldots v_n}}{\vsu}{\tsu},\ldots$,
  then for all types $t_1 \ldots t_n$ there exist values $w_1 \ldots w_n$ such that
  $\steps{\eapp{e}{w_1 \ldots w_n}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$.
\end{cor}

\begin{figure}[t]
\centering
$$
\begin{array}{lclr}
\genWitnessN       & : & \mathsf{Nat} \times e \rightarrow e^{*} & \\
\genWitness{n}{e}  & = & \{ \resolve{\eapp{e}{vs}}{\vsu} \mid \vsu \in \Sigma \} & \\
\quad \mbox{\textbf{where}} &    & & \\
\quad \quad vs     & = & \genArgs{e} & (1) \\
\quad \quad res    & = & \takefn{n}{\evalfn{\eapp{e}{vs}}} & (2) \\
\quad \quad \Sigma & = & \{ \vsu\ \mid \triple{\stuck}{\vsu}{\tsu} \in res\} & (3)
\end{array}
$$
%% \begin{mcode}
%% $\genWitnessN :: (\mathsf{Nat} \times e) \rightarrow 2^{e}$
%% $\genWitness{n}{e} = \{ \resolve{\eapp{e}{vs}}{\vsu} \mid \vsu \in \Sigma \}$
  %% where
   %% $vs     = \genArgs{e}$               -- (1)
   %% $res    = \takefn{n}{\evalfn{\triple{\eapp{e}{vs}}{\emptysu}{\emptysu}}}$  -- (2)
   %% $\Sigma = \{ \vsu\ \mid \triple{\stuck}{\vsu}{\tsu} \in res\}$   -- (3)
%% \end{mcode}
\caption{Generating witnesses.}
\label{fig:algo-gen-witness}
\end{figure}

>>>>

\paragraph{Inputs and Outputs}
%
The problem of generating inputs is undecidable in general.
%
Our witness generation procedure takes two inputs.
%
First, a search bound $k$ which is used to define the \emph{number} of
traces to explore. (We assume, without loss of generality, that all
traces are finite.)
%
Second, the target expression $e$ that contains the type error,
and may be a curried function of multiple arguments.
%
The witness generation procedure returns as output a list of (general)
witness expressions, each of which is of the form $e\ v_1 \ldots v_n$.
%
The \emph{empty} list is returned when no witness can be found after
exploring $n$ traces.

\paragraph{Modeling Semantics}
%
We resolve the non-determinism in the operational semantics
(\S~\ref{sec:semantics}) via the procedure
%
$$
\evalN : e \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu}^{*}
$$
%
Due to the non-determinism introduced by \gensym, a call
$\evalfn{e}$ returns a \emph{list}
of possible results of the form $\triple{v \cup \stuck}{\vsu}{\tsu}$
such that $\steps{e}{\emptysu}{\emptysu}{v \cup \stuck}{\vsu}{\tsu}$.

\paragraph{Currying}
We address the issue of currying by defining a procedure \genArgs{e},
defined in Figure~\ref{fig:algo-gen-args}, that takes as input an
expression $e$ and produces a \emph{saturated} expression of the form
$\eapp{e}{\ehole_1[\thole_1] \ldots \ehole_n[\thole_n]}$ that
\emph{does not} evaluate to a lambda.
%
This is achieved with a simple loop that keeps adding holes to the
target application until evaluating the term yields a non-lambda value.
%
% The helper functions @witness@ and @close@ are used to respectively
% apply the parameters to the target and close the result under the top-level
% let-binders, prior to invoking \hbox{@eval@.}
%
% That is, @mkApps@ creates a nested sequence of applications in
% the usual left-associative style, and @mkLets@ takes a list of
% binders and a body expression, and creates a sequence of nested
% let-binders that close the body expression.

\begin{figure}[t]
% $$
% \begin{array}{lcl}
% \genArgs{e} &\defeq& \begin{cases} \genArgs{\addArg{e}},& \mbox{if } \triple{\efun{x}{e'}}{\vsu}{\tsu},\ldots = \doeval{e} \\
%  e,& \mbox{otherwise} \\
% \end{cases}
% \\
% \addArg{e} &\defeq& \eapp{e}{\vhole{\thole}}, \qquad \mbox{where } \ehole, \thole \mbox{ are fresh }
% \end{array}
% $$
$$
\begin{array}{lclr}
\genArgsN   & : & e \rightarrow e \\
\genArgs{e} & = & \mbox{\textbf{case }} \evalfn{e} \mbox{\textbf{ of}} \\
 \quad \triple{\efun{x}{e}}{\vsu}{\tsu},\ldots &\rightarrow& \genArgs{\eapp{e}{\vhole{\thole}}} & (\ehole, \thole \mbox{ are fresh}) \\
 \quad \_ &\rightarrow& e \\
\end{array}
$$
\caption{Generating a saturated application.}
\label{fig:algo-gen-args}

\end{figure}

\paragraph{Generating Witnesses}
%
Finally, Figure~\ref{fig:algo-gen-witness} summarizes the overall
implementation of our search for witnesses, in the procedure
$\genWitness{k}{e}$, which takes as input a bound $k$ and the
target expression $e$, and returns a list of witness expressions
$\eapp{e}{v_1 \ldots v_n}$ that demonstrate how the input program
gets stuck.
%
The search proceeds as follows.
%
\begin{enumerate}
  \item We invoke $\genArgs{e}$ to produce a \emph{saturated}
        application $e_{sat}$.

  \item We take the first $k$ traces returned by $\evalN$
        on the target $e_{sat}$, and

@@

\begin{cor}{\textbf{[Witness Generation]}}
\label{thm:generation}
  If $$\genWitness{k}{e} = \triple{\eapp{e}{v_1 \ldots v_n}}{\vsu}{\tsu}\ldots$$
  then for all types $t_1 \ldots t_n$ there exist values $w_1 \ldots w_n$ such that
  $\steps{\eapp{e}{w_1 \ldots w_n}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$.
\end{cor}

\begin{figure}[t]
\centering
$$
\begin{array}{lclr}
\genWitnessN       & : & \mathsf{Nat} \times e \rightarrow e^{*} & \\
\genWitness{n}{e}  & = & \{ \resolve{e_{sat}}{\vsu} \mid \vsu \in \Sigma \} & \\
\quad \mbox{\textbf{where}} &    & & \\
\quad \quad e_{sat} & = & \genArgs{e} & (1) \\
\quad \quad res    & = & \takefn{n}{\evalfn{e_{sat}}} & (2) \\
\quad \quad \Sigma & = & \{ \vsu\ \mid \triple{\stuck}{\vsu}{\tsu} \in res\} & (3)
\end{array}
$$
%% \begin{mcode}
%% $\genWitnessN :: (\mathsf{Nat} \times e) \rightarrow 2^{e}$
%% $\genWitness{n}{e} = \{ \resolve{\eapp{e}{vs}}{\vsu} \mid \vsu \in \Sigma \}$
  %% where
   %% $vs     = \genArgs{e}$               -- (1)
   %% $res    = \takefn{n}{\evalfn{\triple{\eapp{e}{vs}}{\emptysu}{\emptysu}}}$  -- (2)
   %% $\Sigma = \{ \vsu\ \mid \triple{\stuck}{\vsu}{\tsu} \in res\}$   -- (3)
%% \end{mcode}
\caption{Generating witnesses.}
\label{fig:algo-gen-witness}
\end{figure}
